FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories

### 2025-02-03 manuscript.Rmd
update 2025-02-03 09:44:07.710923

The first author was inspired by De Groot too,
but they conceive of the empirical cycle in yet another way.
First, notice that the nodes in De Groot's formulation mostly refer to processes.
This invites the question of what the deliverables are in each phase, or in other words: what actually changes when going through the cycle, except the scholar's mind.
In our implementation below, we account for this difference by having the nodes refer to specific deliverables; the edges now refer to processes.
Second, De Groot's strict distinction between processes of observation, induction, and deduction is not widely supported by philosophy of science.
<!-- In our representation, -->
<!-- induction is not a separate phase but a mode of reasoning by which specific observations are generalized into theory. -->
<!-- For example, the refutation of a hypothesized effect, -->
<!-- or the serendipitous observation of some pattern in data, -->
<!-- might be a reason to revise or construct theory. -->
<!-- Induction, incidentally, also occurs within the link from prediction to testing: -->
<!-- in the form of the inductive bias of methods used to perform the test, -->
<!-- and auxiliary assumptions that must be made to address remaining theoretical ambiguities. -->
<!-- # And then for example: I understand Aaron's work about inductive bias to be about the link "prediction -> test;", because the test is not identified without making auxiliary assumptions; if the auxiliary assumption is made based on the data, inductive bias is introduced. -->
<!-- ``` -->
For example, many have argued that observation is value-laden, and as such, involves induction.
The derivation of hypotheses from theory is also not purely deductive,
as auxiliary assumptions are often made (which are, again, an inductive process).
Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well [REF Peikert].
With these alterations, we implement the empirical cycle as follows:

@@

Some have argued that *causal explanations* are a property of good theory [REF Meehl, etc?].
According to Pearl and colleagues,
explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data.
Any formal theory that is explicit about direction of causality could thus be used to guide causal inference,
and could even be integrated into the analysis environment.

>>>>

The first author was inspired by De Groot too,
but they conceive of the empirical cycle in yet another way.
First, notice that the nodes in De Groot's formulation mostly refer to processes.
This invites the question of what the deliverables are in each phase, or in other words: what actually changes when going through the cycle, except the scholar's mind.
In our implementation below, we account for this difference by having the nodes refer to specific deliverables; the edges now refer to processes.
Second, De Groot's distinction between phases of observation, induction, and deduction is not fully congruent with philosophy of science.
<!-- In our representation, -->
<!-- induction is not a separate phase but a mode of reasoning by which specific observations are generalized into theory. -->
<!-- For example, the refutation of a hypothesized effect, -->
<!-- or the serendipitous observation of some pattern in data, -->
<!-- might be a reason to revise or construct theory. -->
<!-- Induction, incidentally, also occurs within the link from prediction to testing: -->
<!-- in the form of the inductive bias of methods used to perform the test, -->
<!-- and auxiliary assumptions that must be made to address remaining theoretical ambiguities. -->
<!-- # And then for example: I understand Aaron's work about inductive bias to be about the link "prediction -> test;", because the test is not identified without making auxiliary assumptions; if the auxiliary assumption is made based on the data, inductive bias is introduced. -->
<!-- ``` -->
For example, many have argued that observation is theory-laden, and as such, involves induction [@brewerTheoryLadennessObservationTheoryLadenness2001].
Deriving hypotheses from theory is also not purely deductive,
as auxiliary assumptions are often made (which, again, involves induction).
Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well [@peikertTransparencyOpenScience2023].
With these alterations, we implement the empirical cycle as follows:

@@

Some have argued that *causal explanations* are a property of good theory [@meehlAppraisingAmendingTheories1990].
According to Pearl and colleagues,
explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data [-@pearlCausalDiagramsEmpirical1995].
Any formal theory that is explicit about direction of causality could thus be used to guide causal inference,
and could even be integrated into the analysis environment.
